Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study

### 2021-06-23 submission/manuscript.rmd
produce word document

---
title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
  **Objective:** Compare risk-based methods for individualizing treatment 
  effects with simulations in the RCT setting. **Study Design and Setting:** We
  predicted absolute benefit based on an available prognostic index (PI) using:
  a model with the PI and a constant relative treatment effect; a model 
  including an interaction of treatment with the PI; 4 quarters of the PI; 
  nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5
  knots); an adaptive model selection method using Akaike’s Information 
  Criterion. Starting from a base case scenario (sample size 4,250, constant 
  odds ratio 0.8, AUC of the PI 0.75), we considered diverse assumptions by
  introducing linear and quadratic interactions of the PI with treatment and 
  varying sample size and discriminative ability of the PI. We evaluated
  performance using root mean squared error, discrimination and calibration for 
  benefit. **Results:** Models including a linear interaction of the PI with
  treatment had adequate performance that was robust under most simulation 
  scenarios. Restricted cubic splines required larger sample sizes and higher 
  AUC of the PI to achieve adequate performance. The adaptive approach performed 
  equivalently to the best-performing method in each scenario. **Conclusion:** 
  Usually, models with just a linear interaction of the PI with treatment
  adequately predict absolute benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

>>>>

---
title: "Continuous risk-based predictive approaches to treatment effect heterogeneity: A simulation study"
abstract: |
  **Objective:** Compare risk-based methods for individualizing treatment 
  effects with simulations in the RCT setting. **Study Design and Setting:** We
  predicted absolute benefit based on an available prognostic index (PI) using:
  a model with the PI and a constant relative treatment effect; a model 
  including an interaction of treatment with the PI; 4 quarters of the PI; 
  nonlinear transformations of the PI (restricted cubic splines with 3, 4 and 5
  knots); an adaptive model selection method using Akaike’s Information 
  Criterion. Starting from a base case scenario (sample size 4,250, constant 
  odds ratio 0.8, AUC of the PI 0.75), we considered diverse assumptions by
  introducing linear and quadratic interactions of the PI with treatment and 
  varying sample size and discriminative ability of the PI. We evaluated
  performance using root mean squared error, discrimination and calibration for 
  benefit. **Results:** Models including a linear interaction of the PI with
  treatment had adequate performance that was robust under most simulation 
  scenarios. Restricted cubic splines required larger sample sizes and higher 
  AUC of the PI to achieve adequate performance. The adaptive approach performed 
  equivalently to the best-performing method in each scenario. **Conclusion:** 
  Usually, models with just a linear interaction of the PI with treatment
  adequately predict absolute benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2: default
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{booktabs}
  - \newcommand\given[1][]{\:#1\vert\:}
  - \date{}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---
